Hydra Battles and Cie

### 2021-02-07 doc/chapter-primrec.tex
start redaction doc on Ackermann function

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}
\begin{Coqsrc}
Inductive PrimRec : nat -> Set :=
  | succFunc : PrimRec 1
  | zeroFunc : PrimRec 0
  | projFunc : forall n m : nat, m < n -> PrimRec n
  | composeFunc :
      forall (n m : nat) (g : PrimRecs n m) (h : PrimRec m), PrimRec n
  | primRecFunc :
      forall (n : nat) (g : PrimRec n) (h : PrimRec (S (S n))), 
      PrimRec (S n)
with PrimRecs : nat -> nat -> Set :=
  | PRnil : forall n : nat, PrimRecs n 0
  | PRcons : forall n m : nat, PrimRec n -> PrimRecs n m -> 
                   PrimRecs n (S m).
\end{Coqsrc}

@@

\index{Projects}
\begin{project}
Prove formally that the Ackermann function is \emph{not} primitive recursive.

\begin{align}
A\,0\,m&=m+1\\
A\,(n+1)\,0 &= A\,n\,1\\
A\,(n+1)\,(m+1)&= A\,n\,(A\,(n+1)\,m)
\end{align}

You may find also  definition of the Ackermann function and a few  lemmas in
 \href{../theories/ordinals/Prelude/Iterates.html}%
{ordinals.Prelude.Iterates}.

\textbf{Hint:} the reader may follow the plan suggested by Bruno Salvy (examen given at the French \'Ecole Polytechnique) (available at
\href{http://www.enseignement.polytechnique.fr/informatique/INF412/uploads/Main/pc-primrec-sujet2014.pdf}{this site}).

\begin{enumerate}
\item Show that, for any $n$, the function $A\,n$ is primitive recursive.
\emph{\textbf{Done.}}

\item Show that for all $n$ and $m$, $A\,n\,m > m$ and  that the functions $A\,n$ are strictly monotonous.
\emph{\textbf{To do.}}

Show also that, for any $n$ and $m$, $A\,n\,m \leq A\,(n+1)\,m$. \emph{\textbf{To do.}}
         Finally, show that for any $n$, $m$ and $k$, 
$A\,k\,(A\,m\,n)\leq A\,(2+\textrm{max}(k,m))\,n$.
\emph{\textbf{To do.}}
\item Show that for any primitive recursive function $f$ of $k$ arguments,
for any $x_1,\dots\,x_k$, $f(x_1,\dots,x_k)\leq A\,n\,(x_1+\dots+x_k)$. \emph{\textbf{To do.}}
\item Show that the Ackermann function is not primitive recursive. \emph{\textbf{To do.}}

\end{enumerate}

\emph{You may start this exercise with
\url{../exercises/primrec/AckermannNotPR.v}.}

@@

\end{project}

>>>>

\label{def:Primrec}
\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}
\begin{Coqsrc}
Inductive PrimRec : nat -> Set :=
  | succFunc : PrimRec 1
  | zeroFunc : PrimRec 0
  | projFunc : forall n m : nat, m < n -> PrimRec n
  | composeFunc :
      forall (n m : nat) (g : PrimRecs n m) (h : PrimRec m), PrimRec n
  | primRecFunc :
      forall (n : nat) (g : PrimRec n) (h : PrimRec (S (S n))), 
      PrimRec (S n)
with PrimRecs : nat -> nat -> Set :=
  | PRnil : forall n : nat, PrimRecs n 0
  | PRcons : forall n m : nat, PrimRec n -> PrimRecs n m -> 
                   PrimRecs n (S m).
\end{Coqsrc}

@@

\section{Proving that a given function is \emph{not} primitive recursive}

The best known of total function which is not primitive recursive is the Ackermann function. We show how to adapt the classic proof (see for instance~\cite{planetmath}) to the constraints of \gallina.

\subsection{The Ackermann function}

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.MoreAck.Ack.html}{Ackermann.MoreAck.Ack}}.

\begin{Coqsrc}
Fixpoint Ack (m:nat) : nat -> nat :=
  match m with
  | 0 => S
  |   S n => fun k =>  iterate (Ack n) (S k) 1
  end.
\end{Coqsrc}

This function is also known under the form of three equations, which we derive immediately from our definition.

\begin{Coqsrc}
Lemma Ack_0 : Ack 0 = S.
Proof refl_equal.

Lemma Ack_S_0 m : Ack (S m) 0 = Ack m 1. 
Proof.  now cbn. Qed.

Lemma Ack_S_S : forall m p,
    Ack (S m) (S p) = Ack m (Ack (S m) p).
Proof.  now cbn. Qed.
\end{Coqsrc}

The order of growth of the Ackermann function w.r.t. its first argument is illustrated by the following equalities.

\begin{Coqsrc}
Lemma Ack_1_n n : Ack 1 n = S (S n).

Lemma Ack_2_n n: Ack 2  n = 2 * n + 3.

Lemma Ack_3_n n: Ack 3 n = exp2 (S (S (S n))) - 3.

Lemma Ack_4_n n: Ack 4 n = hyper_exp2 (S (S (S n))) - 3.
\end{Coqsrc}

Please note that for any given $n$, the unary function
(\texttt{Ack\,$n$} is primitive recursive.

\emph{From \href{../theories/html/hydras.Ackermann.MoreAck.AckNotPR.html}{Ackermann.MoreAck.AckNotPR}}.
\begin{Coqsrc}
Theorem Ackn_IsPR (n: nat) : isPR 1 (Ack n).
\end{Coqsrc}

Ìn order to prove that \texttt{Ack} (considered as a function of two arguments) is not primitive recursive, the usual method consists in two steps:

\begin{enumerate}
\item Prove that for any primitive recursive function $f:\mathbb{N}\rightarrow\mathbb{N}\rightarrow\mathbb{N}$, there exists some natural number $n$ depending on $f$, such that, for any $x$ and $y$, 
$f\,x\,y \leq \texttt{Ack}\,n\,(\textrm{max}\,x\,y)$. Let us say that $f$ is \emph{majorized} [by some \texttt{Ack\,$n$}].
\item Show that \texttt{Ack} fails to satisfy this property.
\end{enumerate}

The second part is the easiest one, let us look at it first, and admit the following lemma.

\begin{Coqsrc}
Lemma PR_2_majorized (f: naryFunc 2)(Hf : isPR 2 f)
  : exists (n:nat), forall x y,  f x y <= Ack n (x + y).
\end{Coqsrc}

Then, assume texttt{Ack} is majorized, and find a contradiction.

\begin{Coqsrc}
Section Impossibility_Proof.

  Hypothesis HAck : isPR 2 Ack.

  Fact Ack_majorized : 
      exists (n:nat), forall x y,  Ack x y <= Ack n (x + y).

  Section Contrad.
    Variable n: nat.
    Hypothesis  Hn :  forall x y,  Ack x y <= Ack n (x + y) .

    Remark R01 : n <> 0.

    Remark R02: n <> 1.

    Remark R03 : max 2 n = n.

    Remark R04 : Ack n (2 * n + 3) = Ack n (Ack 2 n).

    Remark R05 : Ack n (2 * n + 3) <= Ack (n + 2) n.

    Remark R06 : Ack (n + 2) n <= Ack n (2 * n + 2).

    Remark R07 : Ack n (2 * n + 3) <= Ack n (2 * n + 2).

    Remark R08 :  Ack n (2 * n + 2) < Ack n (2 * n + 3).

    Lemma contrad : False.
    Proof.
      generalize R07, R08; intros; lia.
    Qed.

  End Contrad.

  Lemma Ack_not_PR : False.

End Impossibility_Proof.
\end{Coqsrc}

Now it remains to prove Lemma \texttt{PR\_2\_majorized}.
If we look at the inductive definition of primitive recursive functions,\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

\begin{Coqsrc}
Scheme PrimRec_PrimRecs_ind := Induction for PrimRec
  Sort Prop
  with PrimRecs_PrimRec_ind := Induction for PrimRecs 
  Sort Prop.
\end{Coqsrc}

\begin{Coqanswer}
PrimRec_PrimRecs_ind :
forall (P : forall n : nat, PrimRec n -> Prop)
  (P0 : forall n n0 : nat, PrimRecs n n0 -> Prop),
(* successor *)
P 1 succFunc ->

(* zero *)
P 0 zeroFunc ->

* projections *)
(forall (n m : nat) (l : m < n), P n (projFunc n m l)) ->

(* composition *) 
(forall (n m : nat) (g : PrimRecs n m),
      P0 n m g -> forall h : PrimRec m, P m h -> 
      P n (composeFunc n m g h)) ->

(* primitive recursion *)
(forall (n : nat) (g : PrimRec n),
 P n g ->
    forall h : PrimRec (S (S n)), P (S (S n)) h -> 
     P (S n) (primRecFunc n g h)) ->

(* empty list of functions *)
(forall n : nat, P0 n 0 (PRnil n)) ->

(* add a function to a list *)
(forall (n m : nat) (p : PrimRec n),
   P n p -> 
   forall p0 : PrimRecs n m, P0 n m p0 -> 
   P0 n (S m) (PRcons n m p p0)) ->

(* conclusion ! *)
forall (n : nat) (p : PrimRec n), P n p
\end{Coqanswer}

Thus the lemma we will have to prove is, for any primitive recursive function of abritrary arity $n$ is:

\[
   \exists q\in\mathbb{N},\, \forall x_1,\dots,x_n, 
      f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
\]

But dots don't belong to \gallina's syntax! So, we may use \coq's vectors for denoting arbitrary tuples.

%%% ICI

% \item Show that for all $n$ and $m$, $A\,n\,m > m$ and  that the functions $A\,n$ are strictly monotonous.

% Show also that, for any $n$ and $m$, $A\,n\,m \leq A\,(n+1)\,m$. 
%          Finally, show that for any $n$, $m$ and $k$, 
% $A\,k\,(A\,m\,n)\leq A\,(2+\textrm{max}(k,m))\,n$.
% \emph{\textbf{Done.}}
% \item Show that for any primitive recursive function $f$ of $k$ arguments,
% for any $x_1,\dots\,x_k$, $f(x_1,\dots,x_k)\leq A\,n\,(x_1+\dots+x_k)$. \emph{\textbf{Still to do.}}
% \item Show that the Ackermann function is not primitive recursive. \emph{\textbf{Still to do.}}

% \end{enumerate}

@@
